---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

# imdR

<!-- badges: start -->
[![R-CMD-check](https://github.com/Subhradip25/imdR/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/Subhradip25/imdR/actions/workflows/R-CMD-check.yaml)
[![CRAN status](https://www.r-pkg.org/badges/version/imdR)](https://CRAN.R-project.org/package=imdR)
[![License: MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/licenses/MIT)
<!-- badges: end -->

**imdR** is an R package for downloading, processing, and visualizing
India Meteorological Department (IMD) gridded daily meteorological data.
It provides a complete pipeline from raw binary data to climate indices,
trend analysis, and publication-quality maps — all within R.

## What makes imdR different

- **SOI-approved boundaries bundled** — state and district level maps
  with Survey of India shapefiles, ready to use with no extra downloads
- **State and district extraction** — mask any IMD raster to any of
  36 states or 808 districts in one line
- **Complete pipeline** — download → extract → indices → trends → maps
- **No Python required** — pure R implementation with additional
  capabilities beyond existing tools

## Data coverage

| Variable | Resolution | Period |
|---|---|---|
| Rainfall | 0.25 degree | 1901 to present |
| Maximum temperature | 1.0 degree | 1951 to present |
| Minimum temperature | 1.0 degree | 1951 to present |

## Installation

```{r install}
# Install from CRAN
install.packages("imdR")

# Or install the development version from GitHub
install.packages("remotes")
remotes::install_github("Subhradip25/imdR")
```

## Quick start

```{r quickstart}
library(imdR)

file_dir <- tempdir()

# Download 2020 rainfall data
rain2020 <- get_data("rain", 2020, 2020, file_dir)

# Plot a single day — full India map with SOI boundaries
plot_imd(rain2020, "2020-06-28", "rain")

# Zoom to Kerala
plot_imd(rain2020, "2020-06-28", "rain",
         level = "state", name = "Kerala")

# Extract daily time series at Panaji, Goa
goa_ts <- get_point(lat = 15.5, lon = 73.8,
                    variable = "rain",
                    start_yr = 2020, end_yr = 2020,
                    file_dir = file_dir)

# Plot time series with 30-day rolling mean
plot_timeseries(goa_ts, variable = "rain",
                title = "Goa Daily Rainfall 2020")

# Compute rainfall indices for Goa
goa_idx <- compute_rainfall_indices(rain2020,
                                    level    = "state",
                                    name     = "Goa",
                                    file_dir = file_dir)
print(goa_idx)
```

## Functions

| Function | Description |
|---|---|
| `get_data()` | Download and read IMD binary data |
| `open_data()` | Read cached IMD data from disk |
| `get_point()` | Extract daily time series at a point |
| `get_point_all()` | Extract all variables at a point |
| `get_bbox()` | Extract data within a bounding box |
| `extract_by_boundary()` | Mask data to a state or district |
| `list_states()` | List all 36 states and UTs |
| `list_districts()` | List districts, optionally by state |
| `get_boundary()` | Get sf boundary for a state or district |
| `plot_imd()` | Publication-quality daily map |
| `plot_timeseries()` | Daily time series with rolling mean |
| `compute_rainfall_indices()` | 11 rainfall climate indices |
| `compute_temp_indices()` | 13 temperature climate indices |
| `trend_analysis()` | Mann-Kendall test and Sen's slope |

## Climate indices

**Rainfall (11):** dr, d64, d115, rx1day, rx5day, rtwd, sdii,
total, cwd, cdd, pci

**Temperature (13):** mean_tmax, mean_tmin, mean_dtr, txx, txn,
tnx, tnn, su35, su40, tr10, tr25, wsdi, csdi

## Bundled boundaries

imdR includes Survey of India (SOI) approved administrative boundaries
for all 36 Indian states and union territories and 808 districts,
ready to use with no additional downloads.

```{r boundaries}
# List all states
list_states()

# List districts in any state
list_districts("Kerala")

# Get sf boundary object
goa <- get_boundary("state", "Goa")
```

## Data source

IMD gridded data: India Meteorological Department, Pune. https://imdpune.gov.in (accessed periodically; server may be temporarily unavailable)

Boundaries: Survey of India (SOI)

## Citation

```{r citation}
citation("imdR")
```

## Authors

- Subhradip Bhattacharjee  (maintainer)
- Amitava Panja
- Ankita Chakraborty
- Basavareddy
