---
title: "Introduction to imdR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to imdR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

## Overview

`imdR` provides a complete R interface to India Meteorological Department
(IMD) gridded daily meteorological data. It covers:

- **Rainfall** — 0.25 degree resolution, 1901 to present
- **Maximum temperature** — 1.0 degree resolution, 1951 to present
- **Minimum temperature** — 1.0 degree resolution, 1951 to present

The package bundles Survey of India (SOI) approved state and district
boundaries, enabling boundary-aware extraction and publication-quality maps
directly from R.

Two data sources are supported: the **quality-controlled archive**
(`get_data()`, released yearly) and **provisional real-time** daily data
(`get_realtime()`, updated with roughly a one-day lag).

## Installation

```{r install}
# Install from CRAN (once available)
install.packages("imdR")

# Or install the development version from GitHub
# install.packages("remotes")
remotes::install_github("Subhradip25/imdR")
```

```{r load}
library(imdR)
```

## Step 1 — Explore available boundaries

```{r boundaries}
# List all 36 states and union territories
list_states()

# List all districts in a state
list_districts("Goa")
list_districts("Kerala")

# Get the sf boundary object for any state or district
goa_sf      <- get_boundary("state",    "Goa")
north_goa   <- get_boundary("district", "North Goa")
kerala_sf   <- get_boundary("state",    "Kerala")
```

## Step 2 — Download archive data

The archive is the authoritative, quality-controlled dataset, released by
IMD as full-year files.

```{r download}
file_dir <- tempdir()

# Single year — returns a SpatRaster directly
rain2020 <- get_data("rain", 2020, 2020, file_dir)
tmax2020 <- get_data("tmax", 2020, 2020, file_dir)
tmin2020 <- get_data("tmin", 2020, 2020, file_dir)

# Multi-year — returns a named list of SpatRasters (one per year)
# because leap and non-leap years have different layer counts
rain_3yr <- get_data("rain", 2018, 2020, file_dir)

# Read previously downloaded files without re-downloading
rain2020 <- open_data("rain", 2020, 2020, file_dir)
```

## Step 3 — Download real-time (provisional) data

The archive is released yearly with a lag, so the current year is not
available until after it ends. For recent dates — yesterday, this month,
or last month — use `get_realtime()`, which downloads provisional daily
grids from IMD's real-time endpoints.

```{r realtime}
# Yesterday's rainfall across India
y <- Sys.Date() - 1
r_today <- get_realtime("rain", as.character(y), file_dir = file_dir)

# A date range — e.g. this month to date
first <- format(Sys.Date(), "%Y-%m-01")
r_month <- get_realtime("rain", first, as.character(Sys.Date() - 1),
                        file_dir = file_dir)

# Daily rainfall time series at a location (e.g. Agartala, Tripura)
agartala <- get_realtime("rain", "2026-03-01", "2026-06-23",
                         file_dir = file_dir)
df <- to_csv(agartala, lat = 23.83, lon = 91.28)
head(df)
```

**Important caveats:**

- Real-time data is **provisional** and may be revised later by IMD.
  For final analysis, use `get_data()` once the archive year is released.
- Real-time **rainfall** is 0.25 degree (same grid as the archive), but
  real-time **temperature** is 0.5 degree (61 x 61) — different from the
  1.0 degree archive temperature. Do not mix resolutions without resampling.
- Variables: `"rain"`, `"tmax"`, `"tmin"`.

## Step 4 — Visualize

```{r maps}
# Full India map
plot_imd(rain2020, "2020-06-28", "rain")

# Zoom to a state
plot_imd(rain2020, "2020-06-28", "rain",
         level = "state", name = "Kerala",
         save_path = file.path(tempdir(), "rain_Kerala_20200628.png"))

# Zoom to a district
plot_imd(rain2020, "2020-06-28", "rain",
         level = "district", name = "North Goa",
         save_path = file.path(tempdir(), "rain_NorthGoa_20200628.png"))

# Temperature map
plot_imd(tmax2020, "2020-05-20", "tmax",
         save_path = file.path(tempdir(), "tmax_India_20200520.png"))
```

## Step 5 — Point extraction

```{r point}
# Extract daily time series at Panaji, Goa
goa_rain <- get_point(lat = 15.5, lon = 73.8,
                      variable = "rain",
                      start_yr = 2020, end_yr = 2020,
                      file_dir = file_dir)
head(goa_rain)

# Extract all three variables at once (rain + tmax + tmin + DTR)
goa_all <- get_point_all(lat = 15.5, lon = 73.8,
                         start_yr = 2020, end_yr = 2020,
                         file_dir  = file_dir)
head(goa_all)

# Plot the daily time series
plot_timeseries(goa_rain, variable = "rain",
                title = "Goa Daily Rainfall 2020")
```

## Step 6 — Bounding box and boundary extraction

```{r extraction}
# Crop to a bounding box
western_ghats <- get_bbox(
  lat_min = 8, lat_max = 21,
  lon_min = 73, lon_max = 78,
  variable = "rain", start_yr = 2020, end_yr = 2020,
  file_dir = file_dir, format = "netcdf"
)

# Mask to a state boundary
kerala_rain <- extract_by_boundary(
  rain2020, level = "state", name = "Kerala",
  variable = "rain", save = TRUE,
  format = "netcdf", file_dir = file_dir
)

# Mask to a district boundary
north_goa_rain <- extract_by_boundary(
  rain2020, level = "district", name = "North Goa",
  variable = "rain", save = TRUE,
  format = "geotiff", file_dir = file_dir
)
```

## Step 7 — Rainfall climate indices

```{r rainfall-indices}
# Compute 11 indices for full India (2020)
rain_idx <- compute_rainfall_indices(rain2020,
                                     file_dir = file_dir)

# Available indices:
# dr     — rainy days (>= 2.5 mm)
# d64    — heavy precipitation days (>= 64.5 mm)
# d115   — very heavy precipitation days (>= 115.6 mm)
# rx1day — maximum 1-day rainfall
# rx5day — maximum 5-day rainfall
# rtwd   — total rainfall on wet days
# sdii   — simple daily intensity index
# total  — annual total rainfall
# cwd    — consecutive wet days
# cdd    — consecutive dry days
# pci    — precipitation concentration index

# Compute for Goa state only
goa_rain_idx <- compute_rainfall_indices(
  rain2020, level = "state", name = "Goa",
  file_dir = file_dir
)
print(goa_rain_idx)

# Multi-year indices for trend analysis
idx_3yr <- compute_rainfall_indices(rain_3yr,
                                    file_dir = file_dir)
```

## Step 8 — Temperature climate indices

```{r temp-indices}
# Compute 13 temperature indices for full India
temp_idx <- compute_temp_indices(tmax2020, tmin2020,
                                  file_dir = file_dir)

# Available indices:
# mean_tmax — mean daily maximum temperature
# mean_tmin — mean daily minimum temperature
# mean_dtr  — mean diurnal temperature range
# txx       — hottest day (max of tmax)
# txn       — coldest day (min of tmax)
# tnx       — warmest night (max of tmin)
# tnn       — coldest night (min of tmin)
# su35      — summer days (tmax >= 35 C)
# su40      — very hot days (tmax >= 40 C)
# tr10      — cold nights (tmin <= 10 C)
# tr25      — tropical nights (tmin >= 25 C)
# wsdi      — warm spell duration index
# csdi      — cold spell duration index

# Compute for Goa
goa_temp_idx <- compute_temp_indices(
  tmax2020, tmin2020,
  level = "state", name = "Goa",
  file_dir = file_dir
)
print(goa_temp_idx)
```

## Step 9 — Trend analysis

```{r trend}
# Mann-Kendall test + Sen's slope on any index
# Requires at least 3 years of data

# Download a longer time series for meaningful trend
rain_10yr <- get_data("rain", 2011, 2020, file_dir)
idx_10yr  <- compute_rainfall_indices(rain_10yr,
                                       file_dir = file_dir)

# Trend in annual total rainfall
trend_total <- trend_analysis(idx_10yr,
                              index_col = "total",
                              file_dir  = file_dir)

# Trend in rainy days
trend_dr <- trend_analysis(idx_10yr,
                           index_col = "dr",
                           file_dir  = file_dir)

# Region-specific trend — Goa
goa_idx_10yr <- compute_rainfall_indices(
  rain_10yr, level = "state", name = "Goa",
  file_dir = file_dir
)
trend_goa <- trend_analysis(goa_idx_10yr,
                            index_col = "total",
                            name      = "Goa",
                            file_dir  = file_dir)
```

## Output files

All functions that produce files use consistent naming conventions:

| Function | Output file |
|---|---|
| `get_point()` | `rain_15.5N_73.8E_2020_2020.csv` |
| `get_point_all()` | `imd_all_15.5N_73.8E_2020_2020.csv` |
| `get_bbox()` | `rain_8N_21N_73E_78E_2020_2020.nc` |
| `extract_by_boundary()` | `rain_Kerala_2020-01-01_2020-12-31.nc` |
| `compute_rainfall_indices()` | `rainfall_indices_Goa.csv` |
| `compute_temp_indices()` | `temp_indices_Goa.csv` |
| `trend_analysis()` | `trend_total_Goa.csv` + `trend_total_Goa.png` |

## Citation

```{r citation}
citation("imdR")
```

## Data source

IMD gridded data: India Meteorological Department, Pune. https://imdpune.gov.in (accessed periodically; server may be temporarily unavailable)

Boundaries: Survey of India (SOI)
